---
title: "CSH Emergency Presentation"
output: pdf_document
---

```{r libs, include=FALSE}
library(readxl)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
```

```{r import, include=FALSE}
bcm_colorectal_cancer_NK_2_21_25 <- read_excel("bcm_colorectal_cancer.NK.2.21.25.xlsx", 
    col_types = c("numeric", "text", "date", 
        "text", "text", "text", "date", "date", 
        "date", "text", "text", "date", "date"))
```

# Data size

```{r print}
dim(bcm_colorectal_cancer_NK_2_21_25)
```

```{r data prep, include=FALSE}
analytic_data <- bcm_colorectal_cancer_NK_2_21_25 %>% 
  mutate(
    ED_days = as.numeric(Date_of_first_diagnosis - ed_date, units="days"),
    Inp_days = as.numeric(Date_of_first_diagnosis - ip_date, units="days"),
    is_EP = case_when(
      ED_days <= 30 ~ TRUE,
      Inp_days <= 30 ~ TRUE,
      .default = FALSE
    ),
    Age_yrs_at_dx = as.numeric(Date_of_first_diagnosis - birth_dt_tm, units="days") / 365.25,
    Race = as.factor(race_desc_raw)
  )
```




# Univariate

## Times from ED or inpatient to diagnosis

```{r qplots, echo=FALSE, message=FALSE, warning=FALSE}
qplot(analytic_data$ED_days)
qplot(analytic_data$Inp_days)
```

### Distribution of ED $\to$ dx time

```{r summ-ed-days}
summary(analytic_data$ED_days)
```

## Crude EP rate

```{r ep-rate, echo=FALSE}
analytic_data %>%
  count(is_EP) %>% 
  mutate(
    Total = sum(n),
    Percent = round(n / Total * 100, 1)
  ) %>% 
  kable()
```

## Race

```{r race, echo=FALSE}
analytic_data %>% 
  count(Race) %>% 
  arrange(desc(n)) %>% 
  mutate(
    Total = sum(n),
    Percent = round(n / Total * 100, 1)
  ) %>%
  kable()
```




# Bivariate

## Overview of analysis

- Simple proportion of EP divided by all cases (EP rate.)
- Associations with basic demographics
    - age
    - race
    - geography (region)
    - ZIP linked to (if possible):
        - CDC
        - per capita income (quintiles or absolute)

Broadly, same as Cosmos paper. https://pubmed.ncbi.nlm.nih.gov/39394724/

Zimolzak AJ, Khan SP, Singh H, Davila JA. Application of a digital quality measure for cancer diagnosis in Epic Cosmos. *J Am Med Inform Assoc.* 2025;32(1):227--229.

## Age

```{r bivariate-age, echo=FALSE, message=FALSE}
analytic_data %>%
  ggplot(aes(x = Age_yrs_at_dx, fill = is_EP)) +
  geom_histogram()

analytic_data %>% 
  ggplot(aes(x = Age_yrs_at_dx, fill = is_EP)) +
  geom_boxplot(notch = TRUE, outlier.size = 2, outlier.alpha = 0.75, outlier.shape = 1)

t.test(Age_yrs_at_dx ~ is_EP, data = analytic_data)
```

## Race

```{r race-assoc, echo=FALSE}
analytic_data %>% 
  count(is_EP, Race) %>% 
  pivot_wider(names_from = Race, values_from = n) %>% 
  kable()
```
